Computational Prediction of Boundaries of 3D Genomic Domains in Class Imbalance Settings

### 2020-05-08 Manuscript/manuscript.Rmd
added more detail in preciseTAD section of methods

Within this study we used both TAD boundary and loop boundary data derived from the publically available tools ARROWHEAD [@durand2016juicebox] and PEAKACHU [@salameh2019supervised] respectively. ARROWHEAD is part of the Juicer suite of tools for Hi-C data analysis and visualization. To obtain TAD boundaries we applied ARROWHEAD from downloaded primary Hi-C data for the GM12878 and K562 cell lines from Geo GSE63525 [@rao20143d] (libraries HIC001-HIC018 and HIC069-HIC074 for GM12878 and K562 respectively) produced with the MboI restriction enzyme (Supplementary Table 1). The contact matrices were constructed from all read pairs that uniquely align to the genome (MAPQ>0). The default tools for ARROWHEAD were used when extracting the autosomal genomic coordinates (GRCh37/hg19 human genome assembly) of TAD boundaries at 5 kb, 10 kb, 25 kb, 50 kb, and 100 kb resolutions (Additional File 1). For chromatin loop boundaries, we used loops identified by PEAKACHU at 10 kb for the GM12878 and K562 cell line [@salameh2019supervised]. Unique boundaries were considered as the midpoints within the coordinate of each chromatin loop. Chromosome 9 was excluded from all downstream analyses due to the sparsity of the contact matrix at the 5 kb and 10 kb resolutions for the K562 cell line. [??? Can you look up closer? If Arrowhead could call boundaries at 5kb, but not 10kb, it is strange; !!!https://groups.google.com/forum/#!searchin/3d-genomics/chr9$20from$20k562%7Csort:date]. Cell type-specific genomic annotations (BroadHMM chromatin segmentations, histone modifications, and transcription factor binding sites) were obtained from the UCSC Genome Browser Database [@Tyner:2017aa] (GRCh37/hg19 human genome assembly, Supplementary Table 2). [??? Change "cis-regulatory elements (CRE)" to "C; !!!fixed]

@@

In Hi-C, each chromosome is binned into non-overlapping regions of the same length *r*, typically, 5kb and above. The *r* parameter defines the resolution of Hi-C data. Consequently, the location of TAD boundaries is limited by the resolution *r*. We designed a strategy called _shifted binning_ that partitions the genome into regions of the same length *r*, but with middle points corresponding to boundaries defined by the original binning.

@@

Cell type-specific genomic annotations were used to build the predictor space for our TAD boundary prediction framework. Bins were annotated by the number of overlaps with an annotation (*Overlap Count (OC)*), the percent of overlap between the bin and the total width of genomic annotation regions overlapping it (*Overlap Percent (OP)*), the average [??? Is that correct? Or, max?; !!!yes average] signal strength of the corresponding annotation (*Peak Signal Strength*), and the distance in bases from the center of the bin to the center of the nearest genomic annotation region (*Distance*). A log2-transformation of distance was used to account for the skewness present in the distributions of genomic distances (Supplementary Figure 1). Note that model built using a *Peak Signal Strength* feature space where only composed of histone modifications and transcription factor binding sites. That is, they did not include BroadHMM chromatin segmentations.

Chromosome-specific datasets used for modeling were composed of two components: the response vector $Y_{i}$, an indicator variable separating between boundary and non-boundary regions, and the predictor-type specific data matrix, $\textbf{X}_{ij}$, where $i \in \{ 1,2, \cdots,N \}$ ($N$ being the number of chromosome-specific bins) and $j \in \{ 1,2,\cdots,P \}$ ($P$ being the number of genomic annotations). <!--Chromosome-specific datasets were concatenated to form a whole genome dataset. The dataset was then split into a 7:3 ratio for training and testing, respectively.[!!!only comes into play with preciseTAD]-->

@@

Random forest classification models (the `caret` v.6.0 R package [@kuhn2012caret]) were built to compare model performances between combinations of data resolutions, predictor types, and re-sampling techniques. Five-fold cross validation was used to tune the number of annotations, *p*, to consider at each node. The number of trees that were aggregated for each random forest model was set to the default, $n=500$. The performance of each model was compared using Balanced Accuracy (BA), defined as the average of sensitivity and specificity: [??? Correct?;!!!yep]

@@

where TP refers to the number of bins correctly identified as containing a TAD boundary (true positives), FP refers to the number of bins incorrectly identified as containing a TAD boundary (false positives), TN refers to the number of bins correctly identified as not containing a TAD boundary (true negatives), and FN refers to the number of bins incorrectly identified as not containing a TAD boundary (false negatives). Each of these quantities are obtained from the confusion matrix created by validating the model on the testing set. <!--We refer to the complete computational framework used to compare random forest model inputs as *TADRF*.-->

@@

Many genomic annotations, notably architectural proteins, tend to exhibit an extensive pattern of colocalization (correlation) [@van2014insulator]. To suitably reduce the feature space and prevent overfitting, we implemented recursive feature elimination to estimate the optimal number of necessary features to include in the best fitting random forest model. The best performing subset was identified, ranging from 2 to the maximum number of features by power of 2. We then aggregated the predictive importance of the union of the optimal set of features across chromosomes using the mean decrease in node impurity among permuted features in out-of-bag samples to determine the most common and top ranked annotations for predicting TAD boundary regions.

@@

To investigate whether we could alleviate limitations of resolution *r* of Hi-C data that limits the precision of conventional TAD calling algorithms, we applied our optimally built TAD prediction models for predicting TAD boundaries at base pair (bp) resolution. [??? Write how your vector Y and annotation matrices change;!!!added]. At this point, we opted to implement an optimally built whole genome random forest model. That is, a random forest model built on the optimal combination of data resolution (binning), predictor type (feature space), re-sampling technique (class balancing correction), and annotations (most informative genomic elements). That is, $Y$ has length $N$ where $N=n_1+n_2+\cdots+n_{22}$, the total number of bins across each chromosome. The data matrix $\textbf{X}$ has dimensions $N \times p_s$, where $p_s \subset P$, a subset of the most predictive genomic elements.

@@

We first established *predicted TAD boundary regions* (PTBR) as clusters of density-reachable base pairs which have been identified as being a predicted TAD boundary with probability 1 according to the random forest model using *DBSCAN* (version 1.1-5) [@hahsler2019dbscan]. The DBSCAN algorithm takes two parameters, *eps* ($\epsilon$) and *MinPts*. The eps parameter ($\epsilon$) was assigned as the respective data resolution. The heuristic being that base pairs with genomic distances less than resolution should occupy the same designated cluster. The corresponding MinPts parameter was set to 3. [??? This deserves a schematic figure, before Figure 5. And part of it may be in Results;!!!added PTBR to example of Figure 5 A] [??? I would switch the notation. First, you detect REGIONS. They can be adjacent, clustered together, forming AREA. Within this area, you can do pam clustering, and identify POINT;!!!editted, used term "region" instead of "area"] To precisely identify the most representative base pair among each PTBR we performed an extension of partitioning around medoids (PAM), known as CLARA (Clustering Large Applications) for better scaling and more computational efficiency [@reynolds2006clustering; @wei2000empirical]. The midpoint of the corresponding cluster medoid, in this case a particular PTBR, was defined as a *preciseTAD boundary point* (PTBP). <!--Thus, the hierarchical structure of our defined predicted TAD boundary objects were as follows: PTBR > PTBA > PTBP.--> That is, each PTBR is a cluster of base pairs highly probable of being a boundary, with the uniquely identified PTBP being the most representative base pair.

>>>>

Within this study we used both TAD boundary and loop boundary data derived from the publically available tools ARROWHEAD [@durand2016juicebox] and PEAKACHU [@salameh2019supervised] respectively. ARROWHEAD is part of the Juicer suite of tools for Hi-C data analysis and visualization. To obtain TAD boundaries we applied ARROWHEAD using downloaded primary Hi-C data from Geo GSE63525 [@rao20143d] for the GM12878 and K562 cell lines (libraries HIC001-HIC018 and HIC069-HIC074 for GM12878 and K562 respectively) produced with the MboI restriction enzyme (Supplementary Table 1). The contact matrices were constructed from all read pairs that uniquely aligned to the genome (MAPQ>0). The default tools for ARROWHEAD were used when extracting the autosomal genomic coordinates (GRCh37/hg19 human genome assembly) of TAD boundaries at 5 kb, 10 kb, 25 kb, 50 kb, and 100 kb resolutions (Additional File 1). For chromatin loop boundaries, we used loops identified by PEAKACHU at 10 kb for the GM12878 and K562 cell lines [@salameh2019supervised]. Unique boundaries were considered as the midpoints within the coordinate of each chromatin loop. Chromosome 9 was excluded from all downstream analyses due to the sparsity of the contact matrix at the 5 kb and 10 kb resolutions for the K562 cell line. [??? Can you look up closer? If Arrowhead could call boundaries at 5kb, but not 10kb, it is strange; !!!https://groups.google.com/forum/#!searchin/3d-genomics/chr9$20from$20k562%7Csort:date]. Cell line-specific genomic annotations were obtained from the UCSC Genome Browser Database [@Tyner:2017aa] in the form of BroadHMM chromatin segmentations (BroadHMM), histone modifications (HM), and transcription factor binding sites (TFBS) (GRCh37/hg19 human genome assembly, Supplementary Table 2). [??? Change "cis-regulatory elements (CRE)" to "C; !!!fixed]

@@

In Hi-C, each chromosome is binned into non-overlapping regions of the same length *r*, typically 5kb and above. The *r* parameter defines the resolution of Hi-C data. Consequently, the location of TAD boundaries is limited by the resolution *r*. We designed a strategy called _shifted binning_ that partitions the genome into regions of the same length *r*, but with middle points corresponding to boundaries defined by the original binning.

@@

Cell type-specific genomic annotations were used to build the feature space for our boundary prediction framework. Bins were annotated by one of either the number of overlaps with an annotation (*Overlap Count (OC)*), the percent of overlap between the bin and the total width of genomic annotation regions overlapping it (*Overlap Percent (OP)*), the average [??? Is that correct? Or, max?; !!!yes average] signal strength of the corresponding annotation (*Peak Signal Strength*), or the distance in bases from the center of the bin to the center of the nearest genomic annotation region (*Distance*). A $log_2$-transformation of distance was used to account for the skewness present in the distributions of genomic distances (Supplementary Figure 1). It should be noted that models built using a *Peak Signal Strength* feature space were only composed of histone modifications and transcription factor binding sites. That is, they did not include BroadHMM chromatin segmentations.

Chromosome-specific datasets used for modeling were composed of two components: the response vector $Y_{i}$, an indicator variable distinguishing boundary and non-boundary regions, and the predictor-type specific data matrix, $\textbf{X}_{ij}$, where $i \in \{ 1,2, \cdots,N \}$ ($N$ being the number of chromosome-specific bins) and $j \in \{ 1,2,\cdots,P \}$ ($P$ being the number of genomic annotations). <!--Chromosome-specific datasets were concatenated to form a whole genome dataset. The dataset was then split into a 7:3 ratio for training and testing, respectively.[!!!only comes into play with preciseTAD]-->

@@

Random forest (RF) classification models (the `caret` v.6.0 R package [@kuhn2012caret]) were built to compare model performances between combinations of data resolutions, predictor types, and re-sampling techniques. Five-fold cross validation was used to tune the number of annotations, *P*, to consider at each node. The number of trees that were aggregated for each RF model was set to the default, $n=500$. The performance of each model was compared using Balanced Accuracy (BA), defined as the average of sensitivity and specificity: [??? Correct?;!!!yep]

@@

where TP refers to the number of bins correctly identified as containing a boundary (true positives), FP refers to the number of bins incorrectly identified as containing a boundary (false positives), TN refers to the number of bins correctly identified as not containing a boundary (true negatives), and FN refers to the number of bins incorrectly identified as not containing a boundary (false negatives). Each of these quantities are obtained from the confusion matrix created by validating the model on the testing set. <!--We refer to the complete computational framework used to compare random forest model inputs as *TADRF*.-->

@@

Many genomic annotations, notably architectural proteins, tend to exhibit an extensive pattern of colocalization (correlation) [@van2014insulator]. To suitably reduce the feature space and improve computational efficiency, while maintaining optimal performance, we implemented recursive feature elimination. In doing so, we estimated the near optimal number of necessary features to include in the best fitting RF model. The best performing subset was identified, ranging from 2 to the maximum number of features by power of 2. We then aggregated the predictive importance of the union of the optimal set of features across chromosomes using the mean decrease in node impurity among permuted features in out-of-bag samples to determine the most common and top ranked annotations for predicting boundary regions.

@@

To investigate whether we could alleviate resolution limitations of conventional TAD and loop boundary calling algorithms, we applied our RF model to predict boundaries from base pair (bp) resolution data. [??? Write how your vector Y and annotation matrices change;!!!added]. At this point, we opted to implement an optimally built whole genome RF model. That is, a random forest model built on the optimal combination of data resolution (binning), predictor type (feature space), re-sampling technique (class balancing correction), and annotations (most informative genomic elements). Here, $Y$ was of length $N$ where $N=n_1+n_2+\cdots+n_{22}$, the total number of bins across each chromosome. The data matrix $\textbf{X}$ had dimensions $N \times p_s$, where $p_s \subset P$, a subset of the most predictive genomic elements. Given that each chromosome is composed of tens, in some cases hundreds, of millions of base pairs, we used the same whole genome RF model to predict boundaries from chromosome-specific bp resolution data.

@@

First, we extracted the vector $p_{n_{i}}$ that denoted the probability of being a boundary point for each chromosome-specific base pair from our RF model, where $n_{i}$ was the length of chromosome $i$. We then applied *DBSCAN* (version 1.1-5) to the matrix of pairwise distances between base pairs with $p_{n_{i}}=1$. The DBSCAN algorithm takes two parameters, *eps* ($\epsilon$) and *MinPts* [@hahsler2019dbscan]. The eps parameter was assigned as the respective data resolution. The heuristic being that base pairs with genomic distances less than resolution should occupy the same designated cluster. The corresponding MinPts parameter was set to 3. The resulting clusters of density-reachable base pairs identified by DBSCAN were termed *predicted TAD boundary regions* (PTBR) [??? This deserves a schematic figure, before Figure 5. And part of it may be in Results;!!!added PTBR to example of Figure 5 A] [??? I would switch the notation. First, you detect REGIONS. They can be adjacent, clustered together, forming AREA. Within this area, you can do pam clustering, and identify POINT;!!!editted, used term "region" instead of "area"]. To precisely identify the most representative base pair among each PTBR we performed an extension of partitioning around medoids (PAM), known as CLARA (Clustering Large Applications) for better scaling and more computational efficiency [@reynolds2006clustering; @wei2000empirical]. Thusly, the corresponding cluster medoid was defined as a *preciseTAD boundary point* (PTBP), making it the most representative base pair within each clustered PTBR.
